---
title: "Functional programming in R with functionals"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Functional programming in R with functionals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(functionals)
set.seed(1)
```

## Why a functional style

Functional programming builds a computation out of small, referentially
transparent pieces: functions that take input, return output, and do not reach
outside themselves to mutate state. In R this style is natural because functions
are ordinary values -- you can pass them, return them, and store them in lists.

`functionals` is a small toolkit for that style. It gives one consistent family
of *functionals* -- functions that take a function as an argument -- with the
same four arguments everywhere:

| argument      | meaning                                             |
|---------------|----------------------------------------------------|
| `.x` / `.df` / `.l` | the data to iterate over                     |
| `.f`          | the function to apply                               |
| `ncores`      | `NULL`/`1` sequential, `>= 2` parallel              |
| `pb`          | show a completion-driven progress bar               |

and, from version 0.6.0, two more that make loops production-safe:

| argument      | meaning                                             |
|---------------|----------------------------------------------------|
| `.on_error`   | `"stop"` (default), `"pass"`, or `"fill"`           |
| `.seed`       | reproducible per-task RNG streams                   |

Nothing here depends on `rlang` or the tidyverse; the only import is `parallel`.

## The core: `fmap()` and friends

`fmap()` is `lapply()` with the extra arguments above:

```{r}
fmap(1:4, function(x) x^2)
```

The shape-specific mappers save you the `split()` / `as.list()` boilerplate:

```{r}
# one call per data-frame row, row supplied as a named list
fmapr(head(mtcars, 3), function(row) row$mpg / row$cyl)

# one call per column, with the column name available
fmapc(data.frame(a = 1:3, b = 4:6), function(x, nm) sprintf("%s: %.1f", nm, mean(x)))

# one call per group
fmapg(iris, function(d) mean(d$Sepal.Length), by = "Species")

# one call per aligned tuple (list names are matched to .f's arguments)
fmapn(list(x = 1:3, y = 4:6), function(x, y) x + y)
```

`fwalk()` is the side-effect sibling: it runs `.f` for its effect and returns
`.x` invisibly.

```{r}
invisible(fwalk(1:3, function(i) cat("row", i, "\n")))
```

### `floop()` is deprecated

`floop()` did nothing that `fmap()` (collect results) and `fwalk()` (side
effects only) do not. As of 0.6.0 it warns and forwards to them; use those
directly.

## Programming *with* functionals

### Function factories

A function that returns a function pairs naturally with `fmap()`:

```{r}
power_of <- function(k) function(x) x^k
powers   <- fmap(1:4, power_of)          # a list of functions
powers[[3]](2)                            # 2^3
```

### Composition and reduction

`fcompose()` chains single-argument functions right to left; `freduce()` folds a
binary function over a sequence, optionally keeping the running accumulation.

```{r}
standardise <- fcompose(function(x) round(x, 2), scale, as.numeric)
standardise(c(1, 2, 3, 4))

freduce(1:5, `+`)
freduce(1:5, `+`, .accumulate = TRUE)
```

A small pipeline built entirely from functionals -- fit nothing, just transform:

```{r}
pipeline <- fcompose(
  function(x) Filter(function(v) v > 0, x),
  function(x) fmap(x, function(v) v - mean(unlist(x)))
)
str(pipeline(list(-2, 1, 4, 9)))
```

### Parameter sweeps with `fmapn()`

`fmapn()` iterates over aligned lists, so a grid of settings becomes a single
call. `expand.grid()` builds the grid; `fmapn()` walks it.

```{r}
grid <- expand.grid(n = c(10, 50), sd = c(1, 2))
res  <- fmapn(
  list(n = grid$n, sd = grid$sd),
  function(n, sd) c(n = n, sd = sd, se = sd / sqrt(n))
)
do.call(rbind, res)
```

### Safe iteration

With `.on_error = "pass"`, one bad element does not abort the run; the failed
slot holds a `functionals_error` you can inspect afterwards.

```{r}
out <- fmap(list(1, "oops", 3), function(x) x * 10, .on_error = "pass")
vapply(out, function(e) if (inherits(e, "functionals_error")) NA_real_ else e, numeric(1))
```

`.on_error = "fill"` replaces failures with `.fill` directly. The behaviour is
identical whether the run is sequential or parallel.

### Reproducible randomness

`.seed` gives each task an independent L'Ecuyer-CMRG stream, so results do not
depend on `ncores` or the order tasks finish, and your global RNG state is left
untouched.

```{r}
a <- fmap(1:5, function(i) round(rnorm(1), 3), .seed = 42)
b <- fmap(1:5, function(i) round(rnorm(1), 3), .seed = 42, ncores = 1)
identical(a, b)
```

### Parallelism and progress

Set `ncores` to fan work across processes (fork on Unix, PSOCK elsewhere or when
you pass your own `cl`). `pb = TRUE` shows a bar that advances as individual
tasks complete and reports elapsed time and ETA.

```{r eval = FALSE}
slow <- function(x) { Sys.sleep(0.2); x^2 }
fmap(1:20, slow, ncores = 4, pb = TRUE)
```

Objects that cannot survive a fork (external-pointer handles from `torch`,
`xgboost`, database connections, ...) must be created *inside* `.f`, not captured
from the caller. When that is impossible, run sequentially.

## `functionals` and `purrr`

The vocabularies line up almost one to one.

| Task                | `functionals`                          | `purrr`                 | base R                       |
|---------------------|----------------------------------------|-------------------------|------------------------------|
| map                 | `fmap(x, f)`                           | `map(x, f)`             | `lapply(x, f)`               |
| map over N inputs   | `fmapn(list(x, y), f)`                 | `pmap(list(x, y), f)`   | `Map(f, x, y)`               |
| map over rows       | `fmapr(df, f)`                         | `pmap(df, f)`           | `apply(df, 1, f)`            |
| map over columns    | `fmapc(df, f)`                         | `imap(df, f)`           | `Map(f, df, names(df))`      |
| grouped map         | `fmapg(df, f, by = "g")`               | `map(split(df, df$g), f)` | `lapply(split(df, df$g), f)` |
| walk                | `fwalk(x, f)`                          | `walk(x, f)`            | `for (i in x) f(i)`          |
| reduce              | `` freduce(x, `+`) ``                  | `` reduce(x, `+`) ``    | `` Reduce(`+`, x) ``         |
| compose             | `fcompose(f, g)`                       | `compose(f, g)`         | `function(x) f(g(x))`        |
| repeat / simulate   | `frepeat(times = n, expr = f)`         | `map(seq_len(n), ~ f())`| `replicate(n, f())`          |
| parallel + progress | `fmap(x, f, ncores = 4, pb = TRUE)`    | `furrr::future_map()` + `progressr` | `parLapply(cl, x, f)` |

Differences that matter:

* **Dependencies.** `functionals` imports only `parallel`. `purrr` pulls in
  `rlang`, `cli`, `vctrs`, and more.
* **Anonymous functions.** `functionals` takes plain closures only -- no
  `~ .x + 1` formula shorthand. R's native `\(x) x + 1` covers most of that gap.
* **Parallelism is built in.** In `purrr` you switch to `furrr` and set a
  `future` plan; here it is the `ncores` argument.
* **Type-stable output.** `purrr` has `map_dbl()`, `map_chr()`, ...
  `functionals` always returns a list; simplify yourself with `unlist()`,
  `vapply()`, or `do.call(rbind, .)`.

## Performance versus `purrr`

The script `system.file("benchmarks/bench-purrr.R", package = "functionals")`
compares the engines **single-threaded** (`ncores = NULL`, `pb = FALSE`,
`mc.cores = 1`), with an identical closure for every engine, across problem
sizes and two payload weights. The results below are precomputed; re-run the
script to reproduce them on your machine.

```{r bench-table, echo = FALSE}
rds <- system.file("benchmarks/bench-purrr-results.rds", package = "functionals")
if (nzchar(rds)) {
  b <- readRDS(rds)
  r <- b$results
  keep <- r$expression %in% c("functionals::fmap", "purrr::map", "base::lapply")
  wide <- reshape(
    r[keep, c("expression", "n", "payload", "median_ms")],
    idvar = c("n", "payload"), timevar = "expression", direction = "wide"
  )
  names(wide) <- sub("median_ms\\.", "", names(wide))
  wide[order(wide$payload, wide$n), ]
} else {
  message("benchmark rds not installed; run inst/benchmarks/bench-purrr.R")
}
```

```{r bench-note, echo = FALSE, results = "asis"}
if (nzchar(rds)) {
  cat(sprintf(
    "\n_Timings in milliseconds (median). %s, %s; functionals %s, purrr %s._\n",
    b$sysname, b$r_version, b$versions[["functionals"]], b$versions[["purrr"]]
  ))
}
```

The takeaway is deliberately unglamorous: with one thread and a cheap `.f`, all
three are within a small constant factor -- `fmap()` is `lapply()` plus an
argument check, and `purrr::map()` is also compiled C. Raw sequential speed is
not why you reach for `functionals`. The reasons are the *uniform* interface,
built-in `ncores` parallelism, the completion-driven progress bar, and the
`.on_error` / `.seed` controls that the next vignette puts to work in a modeling
pipeline.
